Teaching Debugging Collaboratively---Midway Report

### 2020-05-12 paper/midway.tex
pre-proof

As Michaeli and Romeike point out, there is a lack of research into
the value of teaching debugging in higher education.  None of the
research they found placed much focus on explicitly teaching
debugging.  Chmiel and Loui studied whether students who were provided
with debugging tools and frameworks performed better on tests or spent
less time on assignments than those who were
not\cite{10.1145/971300.971310}.  They weren't able to find conclusive
evidence towards better performance on tests or assignments, though
they did find that students in the treatment group felt more confident
in their debugging abilities.  Unfortunately their study didn't
involve extended explicit teaching of debugging--use of the tools was
voluntary, and variations in the students' individual abilities made
the data difficult to evaluate.\par

@@

\subsection{Overview}

\begin{tikzpicture}
  [node distance = 2em, auto, every node/.style={align=left,
    rectangle, minimum size=2em, rounded corners, fill=blue!20, inner sep=1em},
  ml/.style = {fill=green!30}]

  \node (ll) 
  {\textbf{Low Level:}\\
    RR\\
    gdb
  };
  \node (ml) [ml, right =of ll]
  {\textbf{Mid Level:}\\
    RR\\
    gdb
  };
  \node (hl) [right =of ml]
  {\textbf{High Level:}\\
    Process Vis Tools\\
    Interface to RR/GDB
  };
  \draw[<->] (ll) -- (ml);
  \draw[<->] (ml) -- (hl);
\end{tikzpicture}

\subsection{Low Level}

@@

RR comes with two major limitations: it is only compatible with the
Linux kernel, and it's deterministic recording and replay relies on a
feature that is only found on modern \textit{Intel} x86 CPUs.  For my
purposes the first limitation is fairly irrelevant--I expect students
to be connecting to a central server in order to execute and replay
code, and most systems classes are taught in Linux.  The second
limitation poses more of a problem for instruction.  Because recording
and replay is limited to x86 CPUs, classes cannot use a debugger based
on RR to teach ARM assembly, which is becoming increasingly popular.
Though this isn't a limitation that solutions like the aforementioned
full-VM recorder PANDA face, RR still maintains the advantage of being
a debugger first and foremost.  Students can learn RR, GDB, and
debugging techniques simultaneously.  Platforms that require entire VM
replay come with far greater performance and instructional overhead.
Because of RR's ease of use and tight integration with GDB, I feel it
is the best choice of platform to base a collaborative debugger on.

\subsection{High Level}

@@

\subsection{Mid Level}

@@

>>>>

As Michaeli and Romeike point out, there is a lack of research into
the value of teaching debugging in higher education.  None of the
research they found placed much focus on explicitly teaching
debugging.  Chmiel and Loui studied whether students who were provided
with debugging tools and frameworks performed better on tests or spent
less time on assignments than those who were
not\cite{10.1145/971300.971310}.  They weren't able to find conclusive
evidence towards better performance on tests or assignments, though
they did find that students in the treatment group felt more confident
in their debugging abilities.  Unfortunately their study didn't
involve extended explicit teaching of debugging---use of the tools was
voluntary, and variations in the students' individual abilities made
the data difficult to evaluate.\par

@@

% \subsection{Overview}

\subsection{Preexisting Debugging Tools}

@@

RR comes with two major limitations: it is only compatible with the
Linux kernel, and it's deterministic recording and replay relies on a
feature that is only found on modern \textit{Intel} x86 CPUs.  For my
purposes the first limitation is fairly irrelevant---I expect students
to be connecting to a central server in order to execute and replay
code, and most systems classes are taught in Linux.  The second
limitation poses more of a problem for instruction.  Because recording
and replay is limited to x86 CPUs, classes cannot use a debugger based
on RR to teach ARM assembly, which is becoming increasingly popular.
Though this isn't a limitation that solutions like the aforementioned
full-VM recorder PANDA face, RR still maintains the advantage of being
a debugger first and foremost.  Students can learn RR, GDB, and
debugging techniques simultaneously.  Platforms that require entire VM
replay come with far greater performance and instructional overhead.
Because of RR's ease of use and tight integration with GDB, I feel it
is the best choice of platform to base a collaborative debugger on.

\subsection{User Facing Tools}

@@

% \subsection{Mid Level}

@@

Further research is needed before I start on the user-facing part of
the collaborative debugger.  I feel that it's important to pick and
design visualization tools based on research into what works most
effectively for students.  I also want the visualization tools to be
as general as possible.  Though I plan to use them to visualize x86
programs running in RR, tools like those to visualize registers could
easily be written more generally, allowing them to represent ARM or
MIPS registers.  In order to make the debugger as flexible as possible
I plan on splitting the package into three parts:

\begin{figure}[h]
  \centering
  \begin{tikzpicture}
    [node distance = 2em, auto, every node/.style={align=left,
      rectangle, minimum size=2em, rounded corners, fill=blue!20, inner
      sep=1em}, ml/.style = {fill=green!30}]

    \node (ll) 
    {\textbf{Debugger Facing:}\\
      hooks for specific\\
      RR/GDB output\\
      session management
    };
    \node (ml) [ml, right =of ll]
    {\textbf{Middleware:}\\
      server code\\
      user management\\
      and storage
    };
    \node (hl) [right =of ml]
    {\textbf{User Facing:}\\
      process vis tools\\
      interface to\\
      RR/GDB
    };
    \draw[<->] (ll) -- (ml);
    \draw[<->] (ml) -- (hl);
  \end{tikzpicture}
  \caption{\textbf{Blue:} RR specific software. \textbf{Green:} general collaborative
    debugging management}
\end{figure}

I'm not planning to directly extend RR or GDB to add collaborative
debugging features.  Instead, I plan to write code that will hook into
specific RR/GDB outputs and manage a single RR/GDB session.  This code
will interface with server code that manages multiple users and
coordinates collaborative debugging.  The server will pass on commands
to user facing visualization tools, which will translate output from
the debugger to visualizations and provide an interface to RR/GDB.
Segmenting the code in this way allows the addition of different
debugger backends or different front-end vis tools without changing
the code that handles collaboration.  This makes the debugger
extensible as methods and tools evolve.\par

I plan to continue research and begin working on the software over the
summer.  By the beginning of the Fall semester I hope to have made
significant progress on the interface for a single user to RR/GDB.
